---
title: "Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6
)
```

The `adaptr` package simulates adaptive clinical trials using adaptive
stopping, adaptive arm dropping and/or response-adaptive randomisation.

The package has been developed as part of the
[INCEPT (Intensive Care Platform Trial) project](https://incept.dk/), funded
primarily by a grant from
[Sygeforsikringen "danmark"](https://www.sygeforsikring.dk/).

## Basic example
First, load the package:

```{r}
library(adaptr)
```

### Set up trial
Then, setup a trial with the desired specifications. `adaptr` offers the
general purpose function `setup_trial()`, but here we use the built-in
`setup_trial_binom()` for a trial with a binary, binomially distributed,
undesirable outcome such as mortality (`adaptr` also includes
`setup_trial_norm()` for continuous, normally distributed outcomes).
The example trial specification has three characteristics:

- The allocation probability to each arm cannot be lower than 15% (`min_probs`).
- Equivalence stopping rule: if the simulation yields a 90% probability
(`equivalence_prob`) of treatment differences being < 5 %-points
(`equivalence_diff`), the trial is stopped.
- We soften allocation ratios (`soften_power`) by a constant factor.

```{r}
binom_trial <- setup_trial_binom(
  arms = c("Arm A", "Arm B", "Arm C"),
  true_ys = c(0.25, 0.20, 0.30),
  min_probs = rep(0.15, 3), 
  data_looks = seq(from = 300, to = 2000, by = 100),
  equivalence_prob = 0.9,
  equivalence_diff = 0.05,
  soften_power = 0.5 
) 
```

See `?setup_trial()` for more details on these arguments or
`vignette("Basic-examples", "adaptr")` for basic example trial
specifications and a thorough review of the general settings, and
`vignette("Advanced-example", "adaptr")` for an **advanced** example including
details on how to specify user-written functions for generating outcomes and
posterior draws.

We can print an overview of the trial specification by simply running:

```{r}
binom_trial
```

By default, probabilities are shown with 3 decimals. This can be changed by
explicitly `print`ing the specification with the `prob_digits` arguments, for
example:

```{r}

print(binom_trial, prob_digits = 2)
```

### Simulate a single trial
Remember to define the `seed` to ensure reproducible results. The final trial
results are wrapped here, but when used interactively, you can widen the console
for non-wrapped output:

```{r}
trial_res <- run_trial(binom_trial, seed = 12345)

trial_res
```

Again, we can choose the number of decimals with `print`:

```{r}
print(trial_res, prob_digits = 2)
```

### Simulate multiple trials
Generally, we want to run many simulations using the same trial specification to
assess and compare performance metrics of different trial designs. This is the
job of `run_trials()` (note the final ***s***); again, note the use of a
`base_seed` for reproducible results. Here we run 25 simulations (for tolerable
run-time when producing this example), but in practice you will probably want to
run more simulations.

```{r}
trial_res_mult <- run_trials(binom_trial, n_rep = 25, base_seed = 67890)

trial_res_mult
```

`run_trials()` can run simulations on several CPU cores concurrently: set the
`cores` argument to some number greater than `1` (which is the default value,
resulting in serial/non-parallel processing). As an aside, you can see the
number of available CPU cores by calling `parallel::detectCores()`.

### Get the results of multiple trials
Just printing the resulting object (as the last line above) is not necessarily
what you want. `adaptr` has two convenience functions to get the results of
multiple simulations of the same trial specification, `extract_results()` and
`summary()`; see their documentation for details, e.g. on arm selection in
trials not ending with superiority. The `extract_results()` function yields a
tidy `data.frame` with one simulation per row:

```{r}
extr_res <- extract_results(trial_res_mult)

nrow(extr_res)

head(extr_res)
```

`summary()` (as expected and known from, e.g., regression models in `R`)
summarises the results in a human-friendly manner. Individual key results can be
extracted from the resulting object or printed:

```{r}
res_sum <- summary(trial_res_mult)

print(res_sum, digits = 1)
```

### Visualise trial results
We need non-sparse results for `plot_history` (but *not* for `plot_status`), so
first we re-run `run_trials()`:

```{r}
trial_res_mult <- run_trials(binom_trial, n_rep = 25, base_seed = 67890,
                             sparse = FALSE)
```

Note that the `adaptr` plotting functions require the `ggplot2` package. First,
we plot the overall trial statuses according to the total number of patients
randomised (this does *not* require `sparse = FALSE`):

```{r}
plot_status(trial_res_mult, x_value = "total n")
```

We can also plot the statuses for specific arms. In this example, *Arm A* is
increasingly dropped for inferiority as simulations progress:

```{r}
plot_status(trial_res_mult, x_value = "total n", arm = "Arm A")
```

Next, we plot the history of allocation probabilities at each adaptive
analysis look. Intervals cover the inter-quartile range by default
(`interval_width = 0.5`):

```{r}
plot_history(trial_res_mult)
```

Plotting other summary metrics is possible; see `plot_history()`.

## Citation
If using the package, please consider citing:

```{r}
citation(package = "adaptr")
```
